---
title: "Preprocessing and clustering in scATAC"
output: html_notebook
---
# Install packages 
```{r}
# install EnsDb.Hsapiens.v86
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("EnsDb.Hsapiens.v86")

# install tidyverse
install.packages("tidyverse")

# install ? (Sorry I forgot it...)
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

# install Signac
    # Install bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install()

    # To automatically install Bioconductor dependencies
setRepositories(ind=1:2)

install.packages("Signac")

# Install biovizBase
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biovizBase")
```

```{r}
library(Signac)
library(Seurat)
library(GenomeInfoDb)
library(EnsDb.Hsapiens.v86)
library(patchwork)
set.seed(1234)
library(tidyverse)

DATA_PATH = "../data/scATAC/"
```

# Create a Seurat object
```{r}
source("../scripts/io.R")
```

## For LNCaP cell line
```{r}
counts_LNCaP <- load_scATAC(paste0(DATA_PATH,"LNCaP/"))
# metadata <- read.csv(
#   file = "../vignette_data/atac_v1_lncap_10k_singlecell.csv",
#   header = TRUE,
#   row.names = 1
# )

chrom_assay_LNCaP <- CreateChromatinAssay(
  counts = counts_LNCaP,
  sep = c("_", "_"),
  genome = 'hg38',
  fragments = paste0(DATA_PATH,"LNCaP/",'fragments.tsv.gz'), # needs to be tabix-indexed!
  min.cells = 10,
  min.features = 200
)

lncap <- CreateSeuratObject(
  counts = chrom_assay_LNCaP,
  assay = "peaks_LNCaP"
)
lncap$type = 'DMSO'

granges(lncap)

lncap
```

## For RESA cell line
```{r}
counts_RESA <- load_scATAC(paste0(DATA_PATH,"RESA/"))
# metadata <- read.csv(
#   file = "../vignette_data/atac_v1_lncap_10k_singlecell.csv",
#   header = TRUE,
#   row.names = 1
# )

chrom_assay_RESA <- CreateChromatinAssay(
  counts = counts_RESA,
  sep = c("_", "_"),
  genome = 'hg38',
  fragments = paste0(DATA_PATH,"RESA/",'fragments.tsv.gz'), # needs to be tabix-indexed!
  min.cells = 10,
  min.features = 200
)

resA <- CreateSeuratObject(
  counts = chrom_assay_RESA,
  assay = "peaks_RESA"
)
resA$type = 'ResA'

granges(resA)

```
```{r}
resA
```


# Add gene annotations to allow downstream functions 

## For LNCaP cell line
```{r}
# extract gene annotations from EnsDb
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)

# change to UCSC style since the data was mapped to hg38
seqlevelsStyle(annotations) <- 'UCSC'

# add the gene information to the object
Annotation(lncap) <- annotations
```

## For RESA cell line
```{r}
# add the gene information to the object
Annotation(resA) <- annotations
```

# Computing QC Metrics

## For LNCaP cell line
```{r}
# compute nucleosome signal score per cell
lncap <- NucleosomeSignal(object = lncap)

# compute TSS enrichment score per cell
lncap <- TSSEnrichment(object = lncap, fast = TRUE)

total_fragments <- CountFragments(paste0(DATA_PATH,"LNCaP/",'fragments.tsv.gz'))
sum(is.na(total_fragments))
cb = colnames(lncap) %>% as_tibble %>% rename(CB=value)
total_fragments = inner_join(total_fragments,cb,by="CB")
lncap$fragments <- total_fragments$frequency_count

lncap <- FRiP(
  object = lncap,
  assay = 'peaks_LNCaP',
  total.fragments = 'fragments'
)

lncap$blacklist_fraction <- FractionCountsInRegion(
  object = lncap, 
  assay = 'peaks_LNCaP',
  regions = blacklist_hg38
)

lncap$high.tss <- ifelse(lncap$TSS.enrichment > 2, 'High', 'Low')
TSSPlot(lncap, group.by = 'high.tss') + NoLegend()

VlnPlot(
  object = lncap,
  features = c(
               'nucleosome_signal', 'TSS.enrichment', 'fragments', 'FRiP', 'blacklist_fraction'),
  pt.size = 0.1,
  ncol = 5
)

```

## For RESA cell line
```{r}
# compute nucleosome signal score per cell
resA <- NucleosomeSignal(object = resA)

# compute TSS enrichment score per cell
resA <- TSSEnrichment(object = resA, fast = TRUE)

total_fragments <- CountFragments(paste0(DATA_PATH,"RESA/",'fragments.tsv.gz'))
sum(is.na(total_fragments))
cb_resA = colnames(resA) %>% as_tibble %>% rename(CB=value)
total_fragments_resA = inner_join(total_fragments,cb_resA,by="CB")
resA$fragments <- total_fragments_resA$frequency_count

resA <- FRiP(
  object = resA,
  assay = 'peaks_RESA',
  total.fragments = 'fragments'
)

resA$blacklist_fraction <- FractionCountsInRegion(
  object = resA, 
  assay = 'peaks_RESA',
  regions = blacklist_hg38
)


VlnPlot(
  object = resA,
  features = c(
               'nucleosome_signal', 'TSS.enrichment', 'fragments', 'FRiP', 'blacklist_fraction'),
  pt.size = 0.1,
  ncol = 5
)

```

## Combine 2 datasets
```{r}
combined_scATAC <- merge(x=lncap, y=resA, add.cell.ids=c('DMSO','ResA'))

combined_scATAC
```

```{r}
VlnPlot(
  object = combined_scATAC,
  features = c(
               'nucleosome_signal', 'TSS.enrichment', 'fragments', 'FRiP', 'blacklist_fraction'),
  pt.size = 0.1,
  ncol = 5,
  group.by='type'
)

```

# Subset: remove cells that are outliers for these QC metrics.
## For LNCaP cell line
```{r}
lncap.filt <- subset(
  x = lncap,
  subset = fragments > 2000 &
          fragments < 20000 &
    nucleosome_signal < 9 &
    TSS.enrichment > 2 &
              FRiP > 0.3 &
    blacklist_fraction < 1 
)

lncap.filt

lncap.filt$type = 'DMSO'
# strength of nucleosome-binding pattern: <9
# transcription start site enrichment score as defined by ENCODE: >2
# total number of fragments in peaks: >2000 and < 20000
# fraction of fragments in peaks: >30
# percentage of reads in ENCODE-blacklisted genomic regions: <1
```

## For RESA cell line
```{r}
resA.filt <- subset(
  x = resA,
  subset = fragments > 1000 &
          fragments < 20000 &
    nucleosome_signal < 8 &
    TSS.enrichment > 2 &
              FRiP > 0.4 &
    blacklist_fraction < 1 
)
resA.filt
resA.filt$type = 'ResA'
```

## For merged 
```{r}
combined_scATAC.filt <- merge(x=lncap.filt, y=resA.filt, add.cell.ids=c('DMSO','ResA'))

combined_scATAC.filt
```

```{r}
VlnPlot(
  object = combined_scATAC.filt,
  features = c(
               'nucleosome_signal', 'TSS.enrichment', 'fragments', 'FRiP', 'blacklist_fraction'),
  pt.size = 0.1,
  ncol = 5,
  group.by='type'
)

```


# Normalization and linear dimensional rediction

## For LNCaP cell line
```{r}
lncap.filt <- RunTFIDF(lncap.filt)
lncap.filt <- FindTopFeatures(lncap.filt, min.cutoff = 'q50') #top 50%
lncap.filt <- RunSVD(lncap.filt)

DepthCor(lncap.filt)
```

## For RESA cell line
```{r}
resA.filt <- RunTFIDF(resA.filt)
resA.filt <- FindTopFeatures(resA.filt, min.cutoff = 'q50') #top 50%
resA.filt <- RunSVD(resA.filt)

DepthCor(resA.filt)
```

## For merged 
```{r}
combined_scATAC.filt <- RunTFIDF(combined_scATAC.filt)
combined_scATAC.filt <- FindTopFeatures(combined_scATAC.filt, min.cutoff = 'q50') #top 50%
combined_scATAC.filt <- RunSVD(combined_scATAC.filt)

DepthCor(combined_scATAC.filt) # nothing in the figure
```

# Save 
```{r}
saveRDS(lncap,paste0(DATA_PATH,"LNCaP/seurat_scATAC_lncap_non_qc.rds"))
saveRDS(lncap.filt,paste0(DATA_PATH,"LNCaP/seurat_scATAC_lncap_qc.rds"))

saveRDS(resA,paste0(DATA_PATH,"RESA/seurat_scATAC_ResA_non_qc.rds"))
saveRDS(resA.filt,paste0(DATA_PATH,"RESA/seurat_scATAC_ResA_qc.rds"))

saveRDS(combined_scATAC,paste0(DATA_PATH,"seurat_scATAC_merged_non_qc.rds"))
saveRDS(combined_scATAC.filt,paste0(DATA_PATH,"seurat_scATAC_merged_qc.rds"))
```


<!-- ```{r} -->
<!-- DefaultAssay(lncap) <- "peaks_" -->
<!-- lncap <- RunUMAP(object = lncap, reduction = 'lsi', dims = 2:30) -->
<!-- lncap <- FindNeighbors(object = lncap, reduction = 'lsi', dims = 2:30) -->

<!-- lncap <- FindClusters(object = lncap, verbose = FALSE, algorithm = 3) -->
<!-- DimPlot(object = lncap, label = TRUE) + NoLegend() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- gene.activities <- GeneActivity(lncap) -->
<!-- # add the gene activity matrix to the Seurat object as a new assay and normalize it -->
<!-- lncap[['RNA']] <- CreateAssayObject(counts = gene.activities) -->
<!-- lncap <- NormalizeData( -->
<!--   object = lncap, -->
<!--   assay = 'RNA', -->
<!--   normalization.method = 'LogNormalize', -->
<!--   scale.factor = median(lncap$nCount_RNA) -->
<!-- ) -->

<!-- saveRDS(lncap,paste0(DATA_PATH,"LNCaP/data.rds")) -->
<!-- lncap <- readRDS(paste0(DATA_PATH,"LNCaP/data.rds")) -->
<!-- DefaultAssay(lncap) <- 'RNA' -->

<!-- FeaturePlot( -->
<!--   object = lncap, -->
<!--   features = c('MYC','TP53','GATA2','HOXB13','FOXA1','JUND'), -->
<!--   pt.size = 0.1, -->
<!--   max.cutoff = 'q95', -->
<!--   ncol = 2 -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # change back to working with peaks instead of gene activities -->
<!-- DefaultAssay(lncap) <- 'peaks_' -->
<!-- lncap$seurat_clusters -->
<!-- da_peaks <- FindMarkers( -->
<!--   object = lncap, -->
<!--   ident.1 = "1", -->
<!--   # ident.2 = "2", -->
<!--   min.pct = 0.05, -->
<!--   test.use = 'LR', -->
<!--   # latent.vars = 'peak_region_fragments' -->
<!-- ) -->

<!-- head(da_peaks) -->

<!-- plot1 <- FeaturePlot( -->
<!--   object = lncap, -->
<!--   features = rownames(da_peaks)[1], -->
<!--   pt.size = 0.1 -->
<!-- ) -->
<!-- plot1 -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # set plotting order -->
<!-- # levels(pbmc) <- c("CD4 Naive","CD4 Memory","CD8 Naive","CD8 Effector","DN T","NK CD56bright","NK CD56Dim","pre-B",'pro-B',"pDC","DC","CD14 Mono",'CD16 Mono') -->

<!-- CoveragePlot( -->
<!--   object = lncap, -->
<!--   region = rownames(da_peaks)[4], -->
<!--   extend.upstream = 40000, -->
<!--   extend.downstream = 20000 -->
<!-- ) -->
<!-- ``` -->
